---
title: "ENV 710 Final Project"
author: "Emma Kaufman, Cara Kuuskvere, Jenn McNeill"
date: "2024-04-10"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---
```{r load packages, include=FALSE}

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(lubridate)
library(latex2exp)
library(lubridate)

```


```{r import and wrangle data}

lakes_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_Nutrients_Raw.csv",stringsAsFactors = TRUE) 
do_raw <- read.csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",stringsAsFactors = TRUE)

lakes_raw$sampledate <- mdy(lakes_raw$sampledate)
do_raw$sampledate <- mdy(do_raw$sampledate)

lakes_processed <- lakes_raw %>%
  select(lakename, sampledate, depth, tn_ug, tp_ug, nh34, no23, po4) %>%
  na.omit() %>%
  filter(lakename %in% c("Paul Lake", "West Long Lake", "Peter Lake", "East Long Lake")) %>%
  filter(year(sampledate) >= 1994 & year(sampledate) <= 1999) %>%
  mutate(year = factor(year(sampledate)),
         month = factor(month(sampledate)),
         subsurface = factor(ifelse(depth > 1, "subsurface", "surface")),
         season = factor(ifelse(month == 5, "spring", ifelse(month == 8, "summer", NA)))) %>%
  filter(month %in% c(5,6,7,8)) %>%
  mutate(depth = round(depth * 2) / 2)

lakes_processed$subsurface_indicator <- ifelse(lakes_processed$subsurface == "subsurface", 1, 0)
lakes_processed$summer_indicator <- ifelse(lakes_processed$season == "summer", 1, 0)

lakes_processed <- left_join(lakes_processed, select(do_raw, lakename, sampledate, depth, dissolvedOxygen), by = c("lakename", "sampledate", "depth"))

lake_1994 <- lakes_processed %>% 
  filter(year %in% c(1994)) %>% 
  filter(po4<20) %>% 
  filter(tp_ug<100)

```


```{r import and wrangle data}

# lakes_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_Nutrients_Raw.csv",stringsAsFactors = TRUE) 
# lakes_new <- read_csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
# 
# lakes_raw$sampledate <- mdy(lakes_raw$sampledate)
# 
# lakes_processed <- lakes_raw %>%
#   select(lakename, sampledate, depth, tn_ug, tp_ug, nh34, no23, po4) %>%
#   na.omit() %>%
#   filter(lakename %in% c("Paul Lake", "West Long Lake", "Peter Lake", "East Long Lake")) %>%
#   filter(year(sampledate) >= 1994 & year(sampledate) <= 1999) %>%
#   mutate(year = factor(year(sampledate)),
#          month = factor(month(sampledate)),
#          deep = factor(ifelse(depth > 6, 1, 0))) %>%
#   filter(month %in% c(5,6,7,8))
# 
# lake_1994 <- lakes_processed %>% 
#   filter(year %in% c(1994)) %>% 
#   filter(po4<20) %>% 
#   filter(tp_ug<100)
# 
# lakes_new$sampledate <- mdy(lakes_new$sampledate)
# 
# lakes_chem_processed <- lakes_new %>%
#   select(lakename, sampledate, depth, temperature_C, dissolvedOxygen) %>%
#   na.omit() %>%
#   filter(lakename %in% c("Paul Lake", "West Long Lake", "Peter Lake", "East Long Lake")) %>%
#   filter(year(sampledate) == 1994) %>%
#   mutate(year = factor(year(sampledate)),
#          month = factor(month(sampledate)),
#          deep = factor(ifelse(depth > 6, 1, 0))) %>%
#   filter(month %in% c(5,6,7,8))

```

```{r Season and Phosphorus v DO}
two_season_clean<- lakes_processed %>% 
  na.omit()

two_season_fit <- lm(dissolvedOxygen ~ summer_indicator + tp_ug + summer_indicator*tp_ug,
                     data= two_season_clean)

summary(two_season_fit)


```

a. What is the estimate of $\beta_0$ and what does it represent in terms of DO (dissolved oxygen)?

$\beta_0$ for this fit is 10.22 mg/L, which represents the expected value of DO in spring with zero total phosphorus.

b. What is the estimate of $\beta_1$ and what does it represent in terms of DO?

$\beta_1$ for this fit is -1.52, which represents the difference in DO between spring and summer with zero total phosphorus. In summer the expected dissolved oxygen is 8.7 mg/L. 

c. What is the estimate of $\beta_2$ and what does it represent in terms of DO?

$\beta_2$ for this fit is -0.063 (mg/L)/ug, which represents the expected change in DO for each unit increase in total phosphorus during spring.

d. What is the estimate of $\beta_3$ and what does it represent in terms of DO?

$\beta_3$ for this fit is 0.02 (mg/L)/ug, which represents the adjustment to the slope for each unit increase in total phosphorus during the summer. In the summer the expected change in DO for each unit increase in total phosphorus is -0.045 (mg/L)/ug. 
```{r Season and Phosphorus v DO plot}

Two_Season_P <- ggplot(two_season_clean, aes(x=tp_ug, y=dissolvedOxygen, color=season))+
  geom_point()+
  geom_abline(intercept = coef(two_season_fit)[1],
              slope = coef(two_season_fit)[3],
              color = "thistle3", size = 1)+
  geom_abline(intercept = coef(two_season_fit)[1] + coef(two_season_fit)[2],
              slope = coef(two_season_fit)[3] + coef(two_season_fit)[4],
              color = "darkgoldenrod1", size = 1)+
  scale_color_manual(values = c("spring" = "thistle3", "summer" = "darkgoldenrod1"))+
  scale_x_continuous(breaks = seq(0, 300, by = 50))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Total Phosphorus" ~ (µg)))+
  ylab("dissolved oxygen (mg/L)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "Season")

Two_Season_P
```

```{r looking at residuals for season and phosphorus and MAPE}
plot(two_season_fit)


```



```{r exploratory for appendix, looking at nitrogen and season v DO}

two_season_nitrogen <- lm(dissolvedOxygen ~ summer_indicator + tn_ug + summer_indicator*tn_ug,
                     data= two_season_clean)

 ggplot(two_season_clean, aes(x=tn_ug, y=dissolvedOxygen, color=season))+
  geom_point()+
  geom_abline(intercept = coef(two_season_nitrogen)[1],
              slope = coef(two_season_nitrogen)[3],
              color = "thistle3", size = 1)+
  geom_abline(intercept = coef(two_season_nitrogen)[1] + coef(two_season_nitrogen)[2],
              slope = coef(two_season_nitrogen)[3] + coef(two_season_nitrogen)[4],
              color = "darkgoldenrod1", size = 1)+
  scale_color_manual(values = c("spring" = "thistle3", "summer" = "darkgoldenrod1"))+
  # scale_x_continuous(breaks = seq(0, 300, by = 50))+
  # scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Total Nitrogen" ~ (µg)))+
  ylab("dissolved oxygen (mg/L)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "Season")

```



```{r jenn model}

jenn <- lakes_processed

jenn$subsurface_tp_ug = jenn$tp_ug * jenn$subsurface_indicator
jenn$subsurface_tn_ug = jenn$tn_ug * jenn$subsurface_indicator

jenn_fit_p <- lm(formula = dissolvedOxygen ~ subsurface + tp_ug + subsurface_tp_ug, 
                 data = jenn)

print(jenn_fit_p)

jenn_fit_n <- lm(formula = dissolvedOxygen ~ subsurface + tn_ug + subsurface_tn_ug, 
                 data = jenn)

print(jenn_fit_n)

```

a. What is the estimate of $\beta_0$ and what does it represent in terms of DO?

$\beta_0$ for this fit is 8.65, which represents the expected value of DO in surface water with zero total phosphorus.

b. What is the estimate of $\beta_1$ and what does it represent in terms of DO?

$\beta_1$ for this fit is -0.55, which represents the difference in DO between surface water and subsurface water with zero total phosphorus.

c. What is the estimate of $\beta_2$ and what does it represent in terms of DO?

$\beta_2$ for this fit is -0.0002, which represents the expected change in DO for each unit increase in total phosphorus in surface water.

d. What is the estimate of $\beta_3$ and what does it represent in terms of DO?

$\beta_3$ for this fit is -0.046, which represents the adjustment to the slope for each unit increase in total phosphorus in subsurface water.


```{r jenn plot}

#plot total phosphorus versus dissolved oxygen
ggplot(jenn, aes(x=tp_ug, y=dissolvedOxygen, color=subsurface))+
  geom_point()+
  geom_abline(intercept = coef(jenn_fit_p)[1], 
              slope = coef(jenn_fit_p)[3], 
              color = "steelblue1", size = 1)+
  geom_abline(intercept = coef(jenn_fit_p)[1] + coef(jenn_fit_p)[2], 
              slope = coef(jenn_fit_p)[3] + coef(jenn_fit_p)[4], 
              color = "steelblue4", size = 1)+
  scale_color_manual(values = c("surface" = "steelblue1", "subsurface" = "steelblue4"))+
  scale_x_continuous(breaks = seq(0, 300, by = 50))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab("total phosphorus (micrograms)")+
  ylab("dissolved oxygen (units?)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "indicator")

#plot total nitrogen versus dissolved oxygen
ggplot(jenn, aes(x=tn_ug, y=dissolvedOxygen, color=subsurface))+
  geom_point()+
  geom_abline(intercept = coef(jenn_fit_n)[1], 
              slope = coef(jenn_fit_n)[3], 
              color = "steelblue1", size = 1)+
  geom_abline(intercept = coef(jenn_fit_n)[1] + coef(jenn_fit_n)[2], 
              slope = coef(jenn_fit_n)[3] + coef(jenn_fit_n)[4], 
              color = "steelblue4", size = 1)+
  scale_color_manual(values = c("surface" = "steelblue1", "subsurface" = "steelblue4"))+
  scale_x_continuous(breaks = seq(0, 3500, by = 500))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab("total nitrogen (micrograms)")+
  ylab("dissolved oxygen (units?)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "indicator")

```



## APPENDIX

```{r tp by year with po4}

fit_1 <- lm(formula = tp_ug ~ year + po4, data = lakes_processed)
print(fit_1)

ggplot(lakes_processed, aes(x=po4, y=tp_ug, color=year))+
  geom_point()+
  geom_abline(intercept = coef(fit_1)[1], slope = coef(fit_1)[7], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_1)[1] + coef(fit_1)[2], slope = coef(fit_1)[7], 
              color = "cyan4")+
  geom_abline(intercept = coef(fit_1)[1] + coef(fit_1)[3], slope = coef(fit_1)[7], 
              color = "slateblue")+
  geom_abline(intercept = coef(fit_1)[1] + coef(fit_1)[4], slope = coef(fit_1)[7], 
              color = "tomato3")+
  geom_abline(intercept = coef(fit_1)[1] + coef(fit_1)[5], slope = coef(fit_1)[7], 
              color = "skyblue1")+
  geom_abline(intercept = coef(fit_1)[1] + coef(fit_1)[6], slope = coef(fit_1)[7], 
              color = "limegreen")+
  scale_color_manual(values = c("1994" = "orchid3", 
                                "1995" = "cyan4",
                                "1996" = "slateblue",
                                "1997" = "tomato3",
                                "1998" = "skyblue1",
                                "1999" = "limegreen"))+
  scale_x_continuous(breaks = seq(0, 200, by = 20))+
  scale_y_continuous(breaks = seq(0, 300, by = 50))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
in 1994, 1995, 1996, 1997, 1998, and 1999")+
  theme_light()+
  theme(legend.position="bottom")

```

```{r tp by month with po4}

fit_2 <- lm(formula = tp_ug ~ month + po4, data = lake_1994)
print(fit_2)

ggplot(lake_1994, aes(x=po4, y=tp_ug, color=month))+
  geom_point()+
  geom_abline(intercept = coef(fit_2)[1], slope = coef(fit_2)[5], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[2], slope = coef(fit_2)[5], 
              color = "cyan4")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[3], slope = coef(fit_2)[5], 
              color = "slateblue")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[4], slope = coef(fit_2)[5], 
              color = "tomato3")+
  scale_color_manual(values = c("5" = "orchid3", 
                                "6" = "cyan4",
                                "7" = "slateblue",
                                "8" = "tomato3"))+
  #scale_x_continuous(breaks = seq(0, 200, by = 20))+
  #scale_y_continuous(breaks = seq(0, 300, by = 50))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
in May, June, July, and August")+
  theme_light()+
  theme(legend.position="bottom")

```

```{r tp by deepness binary indicator with po4}

fit_3 <- lm(formula = tp_ug ~ deep + po4, data = lake_1994)
print(fit_3)

ggplot(lake_1994, aes(x=po4, y=tp_ug, color=deep))+
  geom_point()+
  geom_abline(intercept = coef(fit_3)[1], slope = coef(fit_3)[3], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_3)[1] + coef(fit_3)[2], slope = coef(fit_3)[3], 
              color = "cyan4")+
  scale_color_manual(values = c("0" = "orchid3", 
                                "1" = "cyan4"))+
  #scale_x_continuous(breaks = seq(0, 20, by = 20))+
  #scale_y_continuous(breaks = seq(0, 100, by = 50))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
Above and Below 6 ft Water Depth")+
  theme_light()+
  theme(legend.position="bottom")

```

```{r tp by lake with po4}

fit_4 <- lm(formula = tp_ug ~ lakename + po4, data = lake_1994)
print(fit_4)

ggplot(lake_1994, aes(x=po4, y=tp_ug, color=lakename))+
  geom_point()+
  geom_abline(intercept = coef(fit_4)[1], slope = coef(fit_4)[5], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[2], slope = coef(fit_4)[5], 
              color = "cyan4")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[3], slope = coef(fit_4)[5], 
              color = "slateblue")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[4], slope = coef(fit_4)[5], 
              color = "tomato3")+
  scale_color_manual(values = c("East Long Lake" = "orchid3", 
                                "Paul Lake" = "cyan4",
                                "Peter Lake" = "slateblue",
                                "West Long Lake" = "tomato3"))+
  scale_x_continuous(breaks = seq(0, 200, by = 20))+
  scale_y_continuous(breaks = seq(0, 300, by = 50))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
at East Long Lake, Paul Lake, Peter Lake, and West Long Lake")+
  theme_light()+
  theme(legend.position="bottom")

```
