---
title: "ENV 710 Final Project"
author: "Emma Kaufman, Cara Kuuskvere, Jenn McNeill"
date: "2024-04-10"
output: pdf_document
editor_options: 
  chunk_output_type: console
toc: true
---

## Introduction

| Variable Analyzed | Unit of Measure                | Type of Variable           |
|---------------------|-------------------------------|--------------------|
| Lake Name         | Name Factor                    | Character/ Factor          |
| Depth             | Meters below surface           | Numeric ; binary indicator |
| Season            | Spring/ Summer, based on month | Binary indicator           |
| Dissolved Oxygen  | mg/ Liter                      | Numeric                    |
| Phosophorus       | micrograms                     | Numeric                    |

```{r load packages, include=FALSE, message=FALSE, warning=FALSE}

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(lubridate)
library(latex2exp)
library(lubridate)
library(lme4)

```

```{r import and wrangle data, include=FALSE, message=FALSE, warning=FALSE}

lakes_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_Nutrients_Raw.csv",stringsAsFactors = TRUE) 
do_raw <- read.csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",stringsAsFactors = TRUE)

lakes_raw$sampledate <- mdy(lakes_raw$sampledate)
do_raw$sampledate <- mdy(do_raw$sampledate)

lakes_processed <- lakes_raw %>%
  select(lakename, sampledate, depth, tn_ug, tp_ug, nh34, no23, po4) %>%
  na.omit() %>%
  filter(lakename %in% c("Paul Lake", "West Long Lake", "Peter Lake", "East Long Lake")) %>%
  #filter(year(sampledate) >= 1994 & year(sampledate) <= 1999) %>%
  mutate(year = factor(year(sampledate)),
         month = factor(month(sampledate)),
         subsurface = factor(ifelse(depth > 6, "subsurface", "surface")),
         season = factor(ifelse(month == 5, "spring", ifelse(month == 8, "summer", NA)))) %>%
  filter(month %in% c(5,6,7,8)) %>%
  mutate(depth = round(depth * 2) / 2)

lakes_processed$subsurface_indicator <- 
  ifelse(lakes_processed$subsurface == "subsurface", 1, 0)

lakes_processed$summer_indicator <- 
  ifelse(lakes_processed$season == "summer", 1, 0)

lakes_processed <- left_join(lakes_processed, 
                             select(do_raw, lakename, sampledate, depth, dissolvedOxygen), 
                             by = c("lakename", "sampledate", "depth"))

lake_1994 <- lakes_processed %>% 
  filter(year %in% c(1994)) %>% 
  filter(po4<20) %>% 
  filter(tp_ug<100)

```

## MODEL FIT ONE: Dissolved Oxygen by Total Phosphorus with a Seasonal Binary Indicator Variable and a Seasonal Interaction Term

```{r seasonal model, include=FALSE, message=FALSE, warning=FALSE}

two_season_clean <- lakes_processed %>% 
  na.omit()

two_season_fit <- lm(dissolvedOxygen ~ summer_indicator + tp_ug + summer_indicator*tp_ug,
                     data = two_season_clean)

summary(two_season_fit)

```

a.  What is the estimate of $\beta_0$ and what does it represent in terms of DO (dissolved oxygen)?

$\beta_0$ for this fit is 10.14 mg/L, which represents the expected value of DO in spring with zero total phosphorus.

b.  What is the estimate of $\beta_1$ and what does it represent in terms of DO?

$\beta_1$ for this fit is -2.36, which represents the difference in DO between spring and summer with zero total phosphorus. In summer the expected dissolved oxygen is 8.7 mg/L.

c.  What is the estimate of $\beta_2$ and what does it represent in terms of DO?

$\beta_2$ for this fit is -0.067 (mg/L)/ug, which represents the expected change in DO for each unit increase in total phosphorus during spring.

d.  What is the estimate of $\beta_3$ and what does it represent in terms of DO?

$\beta_3$ for this fit is 0.02 (mg/L)/ug, which represents the adjustment to the slope for each unit increase in total phosphorus during the summer. In the summer the expected change in DO for each unit increase in total phosphorus is -0.045 (mg/L)/ug.

| Variable          | Value           | Interpretation                                                                     |
|-------------------|-------------------|------------------------------------|
| Beta0 (Intercept) | 10.14mg/L       | Expected value of DO in spring with 0tp                                            |
| Beta1             | -2.36 mg/L      | Difference in DO between spring and summer with 0 tp                               |
| Beta2             | -0.067 mg/L /ug | Expected change in D0 for each unit increase in tp in the spring                   |
| Beta3             | 0.02 mg/L/ug    | Adjustment to slope for interaction for each unit increase in tp during the summer |

```{r seasonal model holdout and MAPE calculation, include = FALSE, message=FALSE, warning=FALSE}

# Holdout 20% of data for testing, 80% for training

two_season_clean$holdout <- rbinom(n=nrow(two_season_clean),size=1,prob=0.2)
two_season_training <- two_season_clean %>% filter(holdout==0)
two_season_testing <- two_season_clean %>% filter(holdout==1)

two_season_fit_test <- lm(dissolvedOxygen ~ summer_indicator + tp_ug + summer_indicator*tp_ug,
                          data = two_season_training)

print(two_season_fit_test)

two_season_fit_test_beta_0 <- 10.33176
two_season_fit_test_beta_1 <- -2.54724
two_season_fit_test_beta_2 <- -0.07045
two_season_fit_test_beta_3 <- 0.02518

two_season_testing$do_predicted <- two_season_fit_test_beta_0 + 
    two_season_fit_test_beta_1*two_season_testing$summer_indicator +
    two_season_fit_test_beta_2*two_season_testing$tp_ug +
    two_season_fit_test_beta_3*two_season_testing$summer_indicator*two_season_testing$tp_ug

MAPE_seasonal <- sqrt(mean((two_season_testing$dissolvedOxygen - two_season_testing$do_predicted)^2))

print(MAPE_seasonal)

```

```{r seasonal model plot, fig.cap="Figure 1", echo=FALSE, message=FALSE}

ggplot(two_season_clean, aes(x=tp_ug, y=dissolvedOxygen, color=season))+
  geom_point()+
  geom_abline(intercept = coef(two_season_fit)[1],
              slope = coef(two_season_fit)[3],
              color = "thistle3", size = 1)+
  geom_abline(intercept = coef(two_season_fit)[1] + coef(two_season_fit)[2],
              slope = coef(two_season_fit)[3] + coef(two_season_fit)[4],
              color = "darkgoldenrod1", size = 1)+
  scale_color_manual(values = c("spring" = "thistle3", "summer" = "darkgoldenrod1"))+
  scale_x_continuous(breaks = seq(0, 300, by = 50))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Total Phosphorus" ~ (µg)))+
  ylab("Dissolved Oxygen (mg/L)")+
  ggtitle("Dissolved Oxygen vs Total Phosphorus by Season")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "Season")

```

```{r seasonal model residuals, include = FALSE, message=FALSE, warning=FALSE}

plot(two_season_fit)

```

\*\*discuss the poor model fit in results and talk about how we attempted to log transform the x axis and tried to use a subset of the y axis dissolved oxygen, but we still didn't see normal distribution of residuals or constant variance of residuals. reference the plots but include them in the appendix.

## Results Model One

The results of this model are that the expected value of dissolved oxygen across all of the lakes with no phosphorus is 10.22 mg/L in the spring. In the summer, dissolved oxygen levels when there is phosphorus decrease by 1.52 mg/L.

The expected change in DO for each unit increase in total phosphorus during the spring is -0.063 mg/L/ug. In the summer this change increases by 0.02.

These modeled results are displayed with the observed data in [figure \_\_]. These linear models assume that the residuals are normally distributed and have constant variance. The plots that test this are in the appendix, [figures \_\_ and \_\_\_]. These assumptions were not met, as the residuals do not follow a normal distribution (as seen in the qq plot), and the variance is not constant.

We tried to improve our model by normalizing the data with a log transformation of total phosphorus. The results of this are in appendix [figure \_\_]. We still did not see normal distribution of residuals or constant variance of residuals. Additionally, we tried to use a subset of the DO data and fit our model to those data [appendix \_\_\_], but still did not have promising results. We chose to go forward with the original data because it is the most interpretable.

These results show that with increasing concentrations of phosphorus, there is a decrease in dissolved oxygen, and that the impact of phosphorus on DO levels is slightly stronger in the summer. These results are significant (all of the p-values are less than 0.05 for each coefficient), but that doesn't mean they are the best model of these data. The MAPE score of 2.92 shows that there is high error when you compare the fitted and observed data.

## MODEL FIT TWO: Dissolved Oxygen by Total Phosphorus with a Water Depth Binary Indicator Variable and a Water Depth Interaction Term

```{r depth model, include = FALSE, message=FALSE, warning=FALSE}

jenn <- lakes_processed

jenn$subsurface_tp_ug = jenn$tp_ug * jenn$subsurface_indicator
jenn$subsurface_tn_ug = jenn$tn_ug * jenn$subsurface_indicator

jenn_fit_p <- lm(formula = dissolvedOxygen ~ subsurface_indicator + tp_ug + subsurface_tp_ug, 
                 data = jenn)

print(jenn_fit_p)

```

a.  What is the estimate of $\beta_0$ and what does it represent in terms of DO?

$\beta_0$ for this fit is 8.52 mg/L, which represents the expected value of DO in surface water with zero total phosphorus.

b.  What is the estimate of $\beta_1$ and what does it represent in terms of DO?

$\beta_1$ for this fit is -7.67 mg/L, which represents the difference in DO between surface water and subsurface water with zero total phosphorus.

c.  What is the estimate of $\beta_2$ and what does it represent in terms of DO?

$\beta_2$ for this fit is -0.0002 (mg/L)/ug, which represents the expected change in DO for each unit increase in total phosphorus in surface water.

d.  What is the estimate of $\beta_3$ and what does it represent in terms of DO?

$\beta_3$ for this fit is -0.003 (mg/L)/ug, which represents the adjustment to the slope for each unit increase in total phosphorus in subsurface water.

| Variable          | Value       | Interpretation                                                          |
|-------------------|-------------------|------------------------------------|
| Beta0 (Intercept) | 8.52 mg/L   | Expected value of DO in surface water with 0tp                          |
| Beta1             | -7.67 mg/L  | Difference in DO b/w sub- and surface water with 0 tp                   |
| Beta2             | -.0002 mg/L | Expected change in D0 for each unit increase in tp in surface water     |
| Beta3             | -0.003 mg/L | Adjustment to slope for interaction term between tp in subsurface water |

```{r depth model holdout and MAPE calculation, include = FALSE, message=FALSE, warning=FALSE}

# Holdout 20% of data for testing, 80% for training

jenn$holdout <- rbinom(n=nrow(jenn),size=1,prob=0.2)
jenn_training <- jenn %>% filter(holdout==0)

jenn_testing <- jenn %>% filter(holdout==1)

jenn_fit_p_test <- lm(formula = dissolvedOxygen ~ subsurface_indicator + tp_ug + subsurface_tp_ug, data = jenn_training)

print(jenn_fit_p_test)

jenn_fit_p_test_beta_0 <- 8.531495
jenn_fit_p_test_beta_1 <- -7.680194
jenn_fit_p_test_beta_2 <- 0.001925
jenn_fit_p_test_beta_3 <- -0.005985

jenn_testing$do_predicted <- jenn_fit_p_test_beta_0 + 
    jenn_fit_p_test_beta_1*jenn_testing$subsurface_indicator +
    jenn_fit_p_test_beta_2*jenn_testing$tp_ug +
    jenn_fit_p_test_beta_3*jenn_testing$subsurface_tp_ug

MAPE_depth <- sqrt(mean((jenn_testing$dissolvedOxygen - jenn_testing$do_predicted)^2, na.rm = TRUE))

print(MAPE_depth)

```

```{r depth plot, fig.cap = "Figure 2", echo=FALSE, message=FALSE}

#plot total phosphorus versus dissolved oxygen
ggplot(jenn, aes(x=tp_ug, y=dissolvedOxygen, color=subsurface))+
  geom_point()+
  geom_abline(intercept = coef(jenn_fit_p)[1], 
              slope = coef(jenn_fit_p)[3], 
              color = "steelblue1", size = 1)+
  geom_abline(intercept = coef(jenn_fit_p)[1] + coef(jenn_fit_p)[2], 
              slope = coef(jenn_fit_p)[3] + coef(jenn_fit_p)[4], 
              color = "steelblue4", size = 1)+
  scale_color_manual(values = c("surface" = "steelblue1", "subsurface" = "steelblue4"))+
  scale_x_continuous(breaks = seq(0, 300, by = 50))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Total Phosphorus" ~ (µg)))+
  ylab("Dissolved Oxygen (mg/L)")+
  ggtitle("Dissolved Oxygen vs Total Phosphorus by Water Depth")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "depth")

```

```{r depth model residuals, include = FALSE, message=FALSE, warning=FALSE}

plot(jenn_fit_p)

```

## MODEL FIT THREE: Dissolved Oxygen by Total Phosphorus with a Hierarchical Linear Mixed Effects Lake Variable

```{r hierarchical model, include = FALSE, message=FALSE, warning=FALSE}

# Fit hierarchical linear mixed-effects model
fit <- lmer(dissolvedOxygen ~ tp_ug + (1|lakename), data = lakes_processed)

# Summary of the model
summary(fit)

# Print the model
print(fit)

# Extract the random effects
random_effects <- fit@u

```

a.  $\beta_0$ (Intercept): 8.68 mg/L

This represents the expected value of dissolved oxygen with zero total phosphorus.

b.  $\beta_1$ (tp_ug): -0.055 mg/L

This represents the change in dissolved oxygen for each unit increase in total phosphorus in surface water.

| Variable          | Value      | Interpretation                                          |
|-------------------|-------------------|------------------------------------|
| Beta0 (Intercept) | 8.68 mg/L  | Expected value of D0 with zero phosphorus               |
| Beta1 (tp_ug)     | -0.55 mg/L | Change in dissolved oxygen for each unit increase in tp |

```{r hierarchical model holdout and MAPE calculation, include = FALSE, message=FALSE, warning=FALSE}

# Holdout 20% of data for testing, 80% for training
lakes_processed$holdout <- rbinom(n=nrow(lakes_processed),size=1,prob=0.2)
lakes_training <- lakes_processed %>% filter(holdout==0)
lakes_testing <- lakes_processed %>% filter(holdout==1)
fit_test <- lmer(dissolvedOxygen ~ tp_ug + (1|lakename), data = lakes_training)

print(fit_test)

fit_test_beta_0 <- 8.70408
fit_test_beta_1 <- -0.05471

lakes_testing$do_predicted <- fit_test_beta_0 + 
    fit_test_beta_1*lakes_testing$tp_ug

MAPE_hierarchical <- sqrt(mean((lakes_testing$dissolvedOxygen - lakes_testing$do_predicted)^2, na.rm = TRUE))

print(MAPE_hierarchical)

```

```{r hierarchical plot, fig.cap = "Figure 3", echo=FALSE, message=FALSE}

ggplot(lakes_processed, aes(x = tp_ug, y = dissolvedOxygen, color = lakename)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +  # Add a linear regression line
  labs(x = "Total Phosphorus (µg)", y = "Dissolved Oxygen (mg/L)", color = "Lake Name") +
  scale_x_continuous(breaks = seq(0, 300, by = 50))+
  scale_y_continuous(breaks = seq(0, 20, by = 5))+
  theme_light() +
  ggtitle("Dissolved Oxygen vs Total Phosphorus by Lake") +
  theme(legend.position = "right")

```

```{r hierarchical model residuals, include = FALSE, message=FALSE, warning=FALSE}

# Residual Plot (Residuals vs Fitted)
residuals <- resid(fit)
fitted_values <- fitted(fit)

residual_plot <- data.frame(residuals = residuals, fitted_values = fitted_values)

ggplot(residual_plot, aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  ggtitle("Residuals vs Fitted")

# QQ Plot (Normal Q-Q plot)
qq_plot <- qqnorm(residuals, main = "Normal Q-Q Plot")

# Random Effects Plot (Dot plot of random effects)
random_effects <- ranef(fit)$lakename

random_effects_plot <- data.frame(lakename = rownames(random_effects), random_effect = random_effects[, 1])

```

## Model Three Results

This model evaluates dissolved oxygen (dependent) based upon total phosphorus (independent) in each lake in our subset. The hierarchical model evaluates these two variables, with consideration that each lake may have introduce differences in the relationship between the two, whose impacts should not be reflected in the pure randomness between the two variables but can be generalized between all lakes. Like the previous two models, linearity is assumed in the data, along with independence due to the nature of the data collection process. Normality of residuals and homoscedasticity are assumptions made of successful models, however our diagnostic plots demonstrate that this does not

Based on the full dataset, the fixed estimate for the total phosphorus in micrograms is approximately -0.056, and for the training dataset it is approximately -0.054. Given the fixed estimate based on the full dataset, this suggests that for every one-unit increase in total phosphorus, dissolved oxygen decreases by approximately 0.056 units.

## Analysis and Conclusion

| Model                          | MAPE |
|--------------------------------|------|
| Depth Model                    | 2.01 |
| Seasonal Model                 | 2.93 |
| Hierarchical Model by Lakename | 3.07 |

While each of our fits demonstrate slight relationships between the variables investigated, the residual analysis of each indicates that error is not randomly distributed in every model. Further, the plots show two main "clouds" of data, with clusters of points at both high dissolved oxygen and low phosphorus, and consistently low (near zero) dissolved oxygen at the full range of phosphorus levels. This, along with other efforts seen in the appendix aimed at transforming or separating data based on different indicators, failed to separate these two clouds of data points beyond manually sub-setting the data based on dissolved oxygen and phosphorus values.

The depth model has the lowest MAPE value, indicating that this fit is the best of our three we evaluate. This aligns with expectations, as depth should have a larger effect on the quantity of dissolved oxygen due to gas exchange at the surface and mixing. While the depth model is best of the three we evaluate, there are also distinct difference between spring (may) and summer (august) data as well as differecnes between each lake in the relationship between phosphorus and dissolved oxygen.

Scientific evidence demonstrates increases in phosphorus lead to decreases in dissolved oxygen due to eutrophication events in bodies of water. However, this data, along with our research, does not indicate an exact relationship. The two clusters of data may suggest that beyond a certain quantity of phosphorus, or perhaps after a threshold required for a eutrophication event, dissolved oxygen drops to near zero. Thus, while we were unable to uncover quality models to predict the outcome of the quantity of dissolved oxygen based upon our variables, we were able to see this slight negative relationship and critical quantity at which dissolved oxygen largely does not exist beyond near zero values at high quantities of phosophorus.

## APPENDIX

# Model One Residuals

```{r model one residuals, echo = FALSE, message=FALSE, warning=FALSE}

plot(two_season_fit)

```

# Model Two Residuals

```{r model two residuals, echo=FALSE, message=FALSE, warning=FALSE}

plot(jenn_fit_p)

```

# Model Three Residuals

```{r model three residuals, echo = FALSE, message=FALSE, warning=FALSE}

# Residual Plot (Residuals vs Fitted)
residuals <- resid(fit)
fitted_values <- fitted(fit)

residual_plot <- data.frame(residuals = residuals, fitted_values = fitted_values)

ggplot(residual_plot, aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  ggtitle("Residuals vs Fitted")

# QQ Plot (Normal Q-Q plot)
qq_plot <- qqnorm(residuals, main = "Normal Q-Q Plot")

# Random Effects Plot (Dot plot of random effects)
random_effects <- ranef(fit)$lakename

random_effects_plot <- data.frame(lakename = rownames(random_effects), random_effect = random_effects[, 1])

```

```{r exploratory plots, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(lakes_processed)+
  geom_point(aes(y=dissolvedOxygen,x=subsurface_indicator))+
  theme_light()

ggplot(lakes_processed)+
  geom_point(aes(y=dissolvedOxygen,x=season))+
  theme_light()
  
```

```{r tp by month with po4, echo=FALSE, message=FALSE, warning=FALSE}

fit_2 <- lm(formula = tp_ug ~ month + po4, data = lake_1994)
print(fit_2)

ggplot(lake_1994, aes(x=po4, y=tp_ug, color=month))+
  geom_point()+
  geom_abline(intercept = coef(fit_2)[1], slope = coef(fit_2)[5], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[2], slope = coef(fit_2)[5], 
              color = "cyan4")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[3], slope = coef(fit_2)[5], 
              color = "slateblue")+
  geom_abline(intercept = coef(fit_2)[1] + coef(fit_2)[4], slope = coef(fit_2)[5], 
              color = "tomato3")+
  scale_color_manual(values = c("5" = "orchid3", 
                                "6" = "cyan4",
                                "7" = "slateblue",
                                "8" = "tomato3"))+
  scale_x_continuous(breaks = seq(0, 20, by = 5))+
  scale_y_continuous(breaks = seq(0, 100, by = 20))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
in May, June, July, and August")+
  theme_light()+
  theme(legend.position="bottom")

```

```{r tp by lake with po4, echo=FALSE, message=FALSE, warning=FALSE}

fit_4 <- lm(formula = tp_ug ~ lakename + po4, data = lake_1994)
print(fit_4)

ggplot(lake_1994, aes(x=po4, y=tp_ug, color=lakename))+
  geom_point()+
  geom_abline(intercept = coef(fit_4)[1], slope = coef(fit_4)[5], 
              color = "orchid3")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[2], slope = coef(fit_4)[5], 
              color = "cyan4")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[3], slope = coef(fit_4)[5], 
              color = "slateblue")+
  geom_abline(intercept = coef(fit_4)[1] + coef(fit_4)[4], slope = coef(fit_4)[5], 
              color = "tomato3")+
  scale_color_manual(values = c("East Long Lake" = "orchid3", 
                                "Paul Lake" = "cyan4",
                                "Peter Lake" = "slateblue",
                                "West Long Lake" = "tomato3"))+
  scale_x_continuous(breaks = seq(0, 20, by = 5))+
  scale_y_continuous(breaks = seq(0, 100, by = 20))+
  ylab(expression("Total Phosphorus" ~ (µg))) +
  xlab(expression("Phosphate" ~ (µg))) +
  ggtitle("Change in Total Phosphorus in Samples Taken 
at East Long Lake, Paul Lake, Peter Lake, and West Long Lake")+
  theme_light()+
  theme(legend.position="bottom")

```

```{r exploratory for appendix looking at nitrogen and season v DO, echo=FALSE, message=FALSE, warning=FALSE}

two_season_nitrogen <- lm(dissolvedOxygen ~ summer_indicator + tn_ug + summer_indicator*tn_ug,
                     data= two_season_clean)

 ggplot(two_season_clean, aes(x=tn_ug, y=dissolvedOxygen, color=season))+
  geom_point()+
  geom_abline(intercept = coef(two_season_nitrogen)[1],
              slope = coef(two_season_nitrogen)[3],
              color = "thistle3", size = 1)+
  geom_abline(intercept = coef(two_season_nitrogen)[1] + coef(two_season_nitrogen)[2],
              slope = coef(two_season_nitrogen)[3] + coef(two_season_nitrogen)[4],
              color = "darkgoldenrod1", size = 1)+
  scale_color_manual(values = c("spring" = "thistle3", "summer" = "darkgoldenrod1"))+
  # scale_x_continuous(breaks = seq(0, 300, by = 50))+
  # scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Total Nitrogen" ~ (µg)))+
  ylab("dissolved oxygen (mg/L)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "Season")

```

```{r depth model with nitrogen, echo=FALSE, message=FALSE, warning=FALSE}

jenn_fit_n <- lm(formula = dissolvedOxygen ~ subsurface_indicator + tn_ug + subsurface_tn_ug, 
                 data = jenn)

print(jenn_fit_n)

#plot total nitrogen versus dissolved oxygen
ggplot(jenn, aes(x=tn_ug, y=dissolvedOxygen, color=subsurface))+
  geom_point()+
  geom_abline(intercept = coef(jenn_fit_n)[1], 
              slope = coef(jenn_fit_n)[3], 
              color = "steelblue1", size = 1)+
  geom_abline(intercept = coef(jenn_fit_n)[1] + coef(jenn_fit_n)[2], 
              slope = coef(jenn_fit_n)[3] + coef(jenn_fit_n)[4], 
              color = "steelblue4", size = 1)+
  scale_color_manual(values = c("surface" = "steelblue1", "subsurface" = "steelblue4"))+
  scale_x_continuous(breaks = seq(0, 3500, by = 500))+
  scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab("total nitrogen (micrograms)")+
  ylab("dissolved oxygen (units?)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "depth")

```

```{r Heirarchical at surface, echo=FALSE, message=FALSE, warning=FALSE}

lakes_processed_surface <- lakes_processed %>% 
  filter(subsurface_indicator==0)
# Fit hierarchical linear mixed-effects model
fit <- lmer(dissolvedOxygen ~ tp_ug + (1|lakename), data = lakes_processed_surface)

# Summary of the model
summary(fit)

# Print the model
print(fit)

# Extract the random effects
random_effects <- fit@u

# Plotting the relationship between total phosphorus (tp_ug) and dissolved oxygen (dissolvedOxygen) by lakename
ggplot(lakes_processed_surface, aes(x = tp_ug, y = dissolvedOxygen, color = lakename)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +  # Add a linear regression line
  labs(x = "Total Phosphorus (µg)", y = "Dissolved Oxygen (mg/L)", color = "Lake Name") +
  theme_minimal() +
  ggtitle("Dissolved Oxygen vs Total Phosphorus by Lake at Surface") +
  theme(legend.position = "bottom")

```

```{r Heirarchical summer, echo=FALSE, message=FALSE, warning=FALSE}

lakes_processed_summer <- lakes_processed %>% 
  filter(summer_indicator==1)
# Fit hierarchical linear mixed-effects model
fit <- lmer(dissolvedOxygen ~ tp_ug + (1|lakename), data = lakes_processed_summer)

# Summary of the model
summary(fit)

# Print the model
print(fit)

# Extract the random effects
random_effects <- fit@u

# Plotting the relationship between total phosphorus (tp_ug) and dissolved oxygen (dissolvedOxygen) by lakename
ggplot(lakes_processed_summer, aes(x = tp_ug, y = dissolvedOxygen, color = lakename)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +  # Add a linear regression line
  labs(x = "Total Phosphorus (µg)", y = "Dissolved Oxygen (mg/L)", color = "Lake Name") +
  theme_minimal() +
  ggtitle("Dissolved Oxygen vs Total Phosphorus by Lake in Summer") +
  theme(legend.position = "bottom")

```

```{r new wrangling with all years and DO less than 5, echo=FALSE, message=FALSE, warning=FALSE}

lakes_all_years <- lakes_raw %>%
  select(lakename, sampledate, depth, tn_ug, tp_ug, nh34, no23, po4) %>%
  na.omit() %>%
  filter(lakename %in% c("Paul Lake", "West Long Lake", "Peter Lake", "East Long Lake")) %>%
  mutate(year = factor(year(sampledate)),
         month = factor(month(sampledate)),
         subsurface = factor(ifelse(depth > 1, "subsurface", "surface")),
         season = factor(ifelse(month == 5, "spring", ifelse(month == 8, "summer", NA)))) %>%
  filter(month %in% c(5,6,7,8)) %>%
  mutate(depth = round(depth * 2) / 2)

lakes_all_years$subsurface_indicator <- 
  ifelse(lakes_all_years$subsurface == "subsurface", 1, 0)

lakes_all_years$summer_indicator <- 
  ifelse(lakes_all_years$season == "summer", 1, 0)

lakes_all_years <- left_join(lakes_all_years, 
                             select(do_raw, lakename, sampledate, depth, dissolvedOxygen), 
                             by = c("lakename", "sampledate", "depth"))

lakes_all_years <- lakes_all_years %>%
  filter(dissolvedOxygen <= 5)

```

```{r log transformation, echo=FALSE, message=FALSE, warning=FALSE}

two_season_fit_log <- lm(dissolvedOxygen ~ summer_indicator + log(tp_ug) +
                           summer_indicator*tp_ug, data = two_season_clean)

ggplot(two_season_clean, aes(x=log(tp_ug), y=dissolvedOxygen, color=season))+
  geom_point()+
  geom_abline(intercept = coef(two_season_fit_log)[1],
              slope = coef(two_season_fit_log)[3],
              color = "thistle3", size = 1)+
  geom_abline(intercept = coef(two_season_fit_log)[1] + coef(two_season_fit_log)[2],
              slope = coef(two_season_fit_log)[3] + coef(two_season_fit_log)[4],
              color = "darkgoldenrod1", size = 1)+
  scale_color_manual(values = c("spring" = "thistle3", "summer" = "darkgoldenrod1"))+
  #scale_x_continuous(breaks = seq(0, 300, by = 50))+
  #scale_y_continuous(breaks = seq(0, 15, by = 5))+
  xlab(expression("Log Transformed Total Phosphorus" ~ (µg)))+
  ylab("dissolved oxygen (mg/L)")+
  theme_light()+
  theme(legend.position="right")+
  labs(color = "Season")

```
